The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-12-09 draft.tex
version sent to CK and KSL 141209

It is possible to include radiation from a boundary layer (BL) between
the disk and the WD. In \py, the BL is described as
a blackbody with a user-specified effective temperature and
luminosity. In the models presented here, we have followed LK02 in setting
the BL luminosity to zero. However, we have confirmed that the addition of a
BL with $L_{BL} = 0.5 L_{acc}$ and temperatures in the range $80{\rm
kK} \leq T_{BL} \leq 200{\rm kK}$ would not change any of our main
conclusions. 
{\bf CK: Is this true? Do we want to say anything else, e.g. that we'll
talk about this in a future paper? JM: This is true. Anything up to 200kK
still has optical lines. The He II behaviour is not monotnic. CIV drops
gradually and gets a little overionized- but this is a separate topic really.
The issue of clumping and BL and their affect on ionization could be a 
future paper.}

@@

Figure~\ref{wind} shows the physical and ionization structure 
of the benchmark disk wind model. There is an obvious drop-off in density
and temperature with distance away from the disk, so any line
formation process that scales as $n_e^2$ - i.e. recombination and
collisionally excited emission - should be expected to operate
primarily in the dense base of the outflow. Moreover, a comparison of
the rotational and poloidal velocity fields shows that rotation
dominates in the near-disk regime, while outflow dominates further out
in the wind.

@@

The emissivity of these recombination 
features scales as $n_e^2$, meaning that they form almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96). However, as
demonstrated explicitly in section~\ref{modelb}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA\ and (especially) 3934\AA\ 
become quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). Although it has been observed (REF),
this line is somewhat unfamiliar observationally,
but only because it lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

@@

In the UV, the model still produces all the observed lines, 
and deep P-Cygni profiles are produced in the normal resonance lines,
as discussed in section ??. However, the UV spectra also
display what is perhaps the biggest problem with this revised model,
namely the strength of resonance line emission 
at low and intermediate inclinations.
In order to generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model 
is probably somewhat too high to be consistent with UV observations 
of high-state CVs (REF).

@@

\begin{figure} %fullpage
\includegraphics[width=0.8\textwidth]{figures/fig13_eclipse.eps}
\caption{{\sl Top Panel:} In and out of eclipse spectra of the high
inclination NL RW Tri. {\sl Bottom Panel:} In and out of eclipse synthetic
spectra from Model B.
The artificial `absorption' feature just redward of the Balmer jump
is caused for the reasons described in section 5.2.}
\label{rwtricomp}
\end{figure} %fullpage

@@

The similarity between the synthetic and observed spectra is
striking. In particular, the revised model produces strong emission in
all the Balmer lines, with line-to-continuum ratios comparable to
those seen in RW Tri. Moreover, the line-to-continuum contrast
increases during eclipse, as expected for emission produced in a disk
wind. This trend is in line with the observations of RW~Tri, and it
has also been seen in other NLs, including members of the SW~Sex class
(REF) 
% {\bf CK: I'm pretty sure that we know this about UX Uma, e.g. from
%   eclipse mapping studies, and there ought to be mid-eclipse spectra
%   of some additional system}
The revised model also produes a single-peaked \ha\ line, in line with
the observational data.

There are also interesting differences between the revised
model and the RW Tri data set. For example, H$\alpha$ is the only line
that is single-peaked in the model, in contrast to what is seen in the
observations. (However, the relatively low resolution of the RW Tri
spectra makes this comparison difficult.) Moreover, the model exhibits
considerably stronger He~{\sc ii} features than the observations,
which suggests that the overall ionization state of the model is
somewhat too high.

@@

\section{The Dependence of Wind Features on Outflow Parameters}
 Here we could discuss single-peaked emission and trends with parameters.
% \label{modelb}

@@

% \begin{figure} %fullpage
% \mbox{
% \subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_alpha.eps}}
% \quad
% \subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_4686.eps}}   
% }
% \caption{
% Line profiles plotted in veliocity space 
% for \ha\ (left) and \heiiopt\ (right) with varying kinematic 
% properties, computed for an inclination of $80^\circ$.
% The benchmark model and the improved optical
% model described in section 6 are labeled as A and B respectively. 
% The $x$-axis limits correspond to the Keplerian velocity at 
% $4R_{WD}$, the inner edge of the wind.
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
% }
% \label{halpha}
% \end{figure} %fullpage

@@

We have also explored the dependence of the optical (and UV) wind
signatures on some key kinematic wind parameters. Based on this, we
show that recombination emission from outflows with sufficiently high
densities and/or optical depths might 
\renewcommand{\labelitemi}{$\bullet$}
\begin{itemize}
	\item fill in the Balmer absorption edge in the spectrum of
          the accretion, thus accounting for absence of this edge in
          observed CV spectra\citep{KLWB98};
	\item produce the commonly observed single-peaked optical line
          profiles \citep{MC96}
\end{itemize}
\smallskip

{\bf JM: I originally also had a note about the flat Balmer decrement here.
it's potentially interesting and could be included, but would need a discussion
earlier in the text.}

\noindent Based on our modelling, we can also make a number of basic,
testable predictions for wind-formed signatures:
\begin{itemize}
	\item we expect the depth of the Balmer absorption edge in CV
          spectra to decrease with increasing inclination (with the
          Balmer jump possibly switching to emission at the highest
          inclinations, if the flux in the wind-formed recombination
          continuum exceeds that of the foreshortened disk spectrum);
	\item we predict that a strong He~\textsc{ii}~$\lambda3202{\rm
          \AA}$ line should be prominent in all cases where
          He~\textsc{ii}~$\lambda1640{\rm \AA}$ \& \heiiopt\ are seen
          (but note that this prediction is not restricted to lines
          formed in disk winds);
	\item we predict that the Balmer jump should be in emission
          during mid-eclipse in high inclination systems.
\end{itemize}

\smallskip
We have also constructed a revised benchmark model which is designed
to more closely match the optical spectra of high-state CVs. This
optically optimized model produces all the prominent optical lines in
and out of eclipse, and achieves reasonable verisimilitude with the
observed optical spectra of RW Tri. However, this model also has
significant shortcomings. In particular, it predicts
stronger-than-observed He~{\sc ii} lines in the optical region and too
much of a collisionally excited contribution to the UV resonance lines
that

@@

{\bf CK: I just realized that -- apart from the intro -- we never
actually discuss the issue of P Cyg profiles in the optical
lines. These have been observed -- e.g. in Halpha -- but I guess we
never produce these? Do we know why? We probably should at least
comment on this at some point in the text, and possibly even briefly
in the conclusions.}

@@

\subsection*{Acknowledgements}
The work of JHM, NH and CK is supported by the Science and Technology Facilities Council (STFC), 
via studentships and a consolidated grant, respectively. We would like to thank Juan Hernandez Santisteban and Ivan Hubeny 
for useful discussions. We acknowledge the use of the IRIDIS High Performance Computing Facility, 
and associated support services at the University of Southampton, in the completion of this work.

\newpage
\newpage
\bibliography{mybib.bib}

>>>>

It is possible to include radiation from a boundary layer (BL) between
the disk and the WD. In \py, the BL is described as
a blackbody with a user-specified effective temperature and
luminosity. In the models presented here, we have followed LK02 in setting
the BL luminosity to zero. However, we have confirmed that the addition of a
BL with $L_{BL} = 0.5 L_{acc}$ and temperatures in the range $80{\rm
kK} \leq T_{BL} \leq 200{\rm kK}$ would not change any of our main
conclusions.

@@

Figure~\ref{wind} shows the physical and ionization structure 
of the benchmark disk wind model. There is an obvious drop-off in density
and temperature with distance away from the disk, so any line
formation process that scales as $\rho^2$ - i.e. recombination and
collisionally excited emission - should be expected to operate
primarily in the dense base of the outflow. Moreover, a comparison of
the rotational and poloidal velocity fields shows that rotation
dominates in the near-disk regime, while outflow dominates further out
in the wind.

@@

The emissivity of these recombination 
features scales as $\rho^2$, meaning that they form almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96). However, as
demonstrated explicitly in section~\ref{modelb}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA\ and (especially) 3934\AA\ 
become quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). Although it has been observed (REF),
this line is somewhat unfamiliar observationally. 
This is because it lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

In the UV, the model still produces all the observed lines, 
and deep P-Cygni profiles are produced in the normal resonance lines,
as discussed in section ??. However, the UV spectra also
display what is perhaps the biggest problem with this revised model,
namely the strength of resonance line emission 
at low and intermediate inclinations.
In order to generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA\ resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model 
is probably somewhat too high to be consistent with UV observations 
of high-state CVs (REF).

\subsection{Continuum Shape and the Balmer Jump}

@@

It has been noted by \cite{ladous1989, ladous1989b}
that the Balmer jump in high-state CVs would naturally weaken at
high inclinations due to limb darkening effects. 
Although we include a simple limb darkening law which affects 
the {\em total flux} seen at each incliantion, we cannot
include it as an effect on the {\em opacities} in our model.
As a result, the efficiency of filling in the Balmer jump
should really be judged at low and medium inclinations, 
where, although prominent, the recombination continuum does
not overcome the disk atmosphere absorption. 
In addition, this effect 
could mean that any model which successfully fills in the 
jump at low inclinations could lead to a Balmer jump 
in emission at high inclinations.
To understand this phenomenon, a fully self-consistent
radiative transfer calculation of both the disk atmosphere
and connected wind is required.

\subsection{Line Profile Shapes: Producing Single-Peaked Emission}

% Although \ha\ is single-peaked in our model, the majority of the
% lines are double-peaked, particularly  
Figure~\ref{halpha} shows how the H$\alpha$ and 
% He~{\sc ii}~$\lambda 4686$ 
line changes with the kinematics of the wind 
for an inclination of $80^\circ$. The width and strength of He~{\sc ii} 
clearly depend on both $R_v$ and $\alpha$, but the overall line profile 
shape is always double-peaked. This contrasts with the behaviour of 
H$\alpha$, which becomes single-peaked in our densest models. The transition from
double-peaked to single-peaked occurs when the line becomes optically
thick. This reproduces the behaviour predicted by \cite{MC96}, who
showed that the poloidal velocity gradient in the wind can modify the
shape of optically thick emission lines even in the near-disk regime,
where the poloidal velocity itself is much smaller than the rotational
velocity. This has the effect of suppressing the line wings and
can therefore lead to the production of a single-peaked line profile
shape.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in veliocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a transition from double to singl-peaked lines
as the emission region becomes more optically thick.
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

@@

\begin{figure} %fullpage
\includegraphics[width=0.9\textwidth]{figures/fig13_eclipse.eps}
\caption{{\sl Top Panel:} In and out of eclipse spectra of the high
inclination NL RW Tri. {\sl Bottom Panel:} In and out of eclipse synthetic
spectra from Model B.
The artificial `absorption' feature just redward of the Balmer jump
is caused for the reasons described in section 5.2.}
\label{rwtricomp}
\end{figure} %fullpage

@@

The similarity between the synthetic and observed spectra is
striking. In particular, the revised model produces strong emission in
all the Balmer lines, with line-to-continuum ratios comparable to
those seen in RW Tri. Moreover, the line-to-continuum contrast
increases during eclipse, as expected for emission produced in a disk
wind. This trend is in line with the observations of RW~Tri, and it
has also been seen in other NLs, including members of the SW~Sex class
(REF) 
% {\bf CK: I'm pretty sure that we know this about UX Uma, e.g. from
%   eclipse mapping studies, and there ought to be mid-eclipse spectra
%   of some additional system}
% The revised model also produes a single-peaked \ha\ line, in line with
% the observational data.

There are also interesting differences between the revised
model and the RW Tri data set. For example, the model exhibits
considerably stronger He~{\sc ii} features than the observations,
which suggests that the overall ionization state of the model is
somewhat too high. 
As discussed in section 5.3, H$\alpha$ is the only line
that is single-peaked in the model. This is
in contrast to what is seen in the observations,
although the relatively low resolution of the RW Tri
spectra makes this comparison difficult.

@@

% \section{The Dependence of Wind Features on Outflow Parameters}
%  Here we could discuss single-peaked emission and trends with parameters.
% \label{modelb}

@@

@@

We have also constructed a revised benchmark model which is designed
to more closely match the optical spectra of high-state CVs. This
optically optimized model produces all the prominent optical lines in
and out of eclipse, and achieves reasonable verisimilitude with the
observed optical spectra of RW Tri. However, this model also has
significant shortcomings. In particular, it predicts
stronger-than-observed He~{\sc ii} lines in the optical region and too
much of a collisionally excited contribution to the UV resonance lines
that

Based on this, we
show that recombination emission from outflows with sufficiently high
densities and/or optical depths might

\renewcommand{\labelitemi}{$\bullet$}
\begin{itemize}
	\item fill in the Balmer absorption edge in the spectrum of
          the accretion, thus accounting for absence of this edge in
          observed CV spectra \citep{KLWB98};
	\item produce the commonly observed single-peaked optical line
          profiles \citep{MC96}
\end{itemize}
\smallskip

Although we note that limb darkening effects on disk atmosphere opacities
would affect the Balmer jump. We can also make a number of basic,
testable predictions for wind-formed signatures:

% {\bf JM: I originally also had a note about the flat Balmer decrement here.
% it's potentially interesting and could be included, but would need a discussion
% earlier in the text.}

\noindent We can also make a number of basic,
testable predictions for wind-formed signatures:
\begin{itemize}
	\item we expect the depth of the Balmer absorption edge in CV
          spectra to decrease with increasing inclination (with the
          Balmer jump possibly switching to emission at the highest
          inclinations, if the flux in the wind-formed recombination
          continuum exceeds that of the foreshortened disk spectrum);
	\item we predict that a strong He~\textsc{ii}~$\lambda3202{\rm
          \AA}$ line should be prominent in all cases where
          He~\textsc{ii}~$\lambda1640{\rm \AA}$ \& \heiiopt\ are seen
          (but note that this prediction is not restricted to lines
          formed in disk winds);
	\item we predict that the Balmer jump should be in emission
          during mid-eclipse in high inclination systems.
\end{itemize}

\smallskip

@@

% {\bf CK: I just realized that -- apart from the intro -- we never
% actually discuss the issue of P Cyg profiles in the optical
% lines. These have been observed -- e.g. in Halpha -- but I guess we
% never produce these? Do we know why? We probably should at least
% comment on this at some point in the text, and possibly even briefly
% in the conclusions.}

@@

\subsection*{Acknowledgements}
The work of JHM and CK is supported by the Science and Technology Facilities Council (STFC), 
via studentships and a consolidated grant, respectively. We would like to thank Juan Hernandez Santisteban, Samuel Mangham and Ivan Hubeny for useful discussions. 
We acknowledge the use of the IRIDIS High Performance Computing Facility, 
and associated support services at the University of Southampton, in the completion of this work.

\bibliography{mybib.bib}
